Efficient ancestry and mutation simulation with msprime 1.0

### 2021-08-16 paper.tex
Editing pass

\begin{abstract}
Stochastic simulation is a key tool in population genetics,
since the models involved are often analytically intractable
and simulation is usually the only way of obtaining ground-truth data to
evaluate inferences.
Because of this necessity, a large number
of specialised simulation programs have been developed, each
filling a particular niche, but with largely overlapping functionality
and a substantial duplication of effort.
Since its introduction, the \msprime\ simulator has significantly
extended the scope of coalescent simulation, enabling the
simulation and efficient analysis of datasets on a far
larger scale than previously possible.
Here, we introduce \msprime\ version 1.0, which
efficiently implements ancestry and mutation simulations using
the succinct tree sequence data structure and \tskit\ library.
We summarise \msprime's many features, and show that
its performance is excellent, often many times faster
and more memory efficient than specialised alternatives.
These high-performance features have been thoroughly
tested and validated, and built using a collaborative,
open source development model,
which reduces duplication of effort and promotes
software quality via community engagement.
% "Bazaar" development model: https://en.wikipedia.org/wiki/The_Cathedral_and_the_Bazaar
\end{abstract}

@@

The coalescent
process~\citep{kingman1982coalescent,hudson1983testing,tajima1983evolutionary}
models the ancestry of a set of sampled genomes,
providing a mathematical description of the
genealogical tree that relates the samples to one another.
It has proved to be a powerful model,
and is now central to
population genetics~\citep{hudson1990gene,hein2004gene,wakely2008coalescent}.
The coalescent is a very efficient framework for population genetic
simulation, because it allows us to simulate the genetic ancestry for
a sample from a Wright-Fisher population, without explicitly representing
the population in memory or stepping through the generations.
Indeed, \cite{hudson1983testing} independently derived the
% Anyone disagree here? That's the impression I get from the paper.
coalescent \emph{in order to} efficiently simulate data,
and he used these simulations to characterise an analytically
intractable distribution.
This efficiency and the great utility of simulations for a wide
range of purposes has led to dozens of different tools
being developed over the decades~\citep{carvajal2008simulation,liu2008survey,
arenas2012simulation,yuan2012overview,hoban2012computer,yang2014critical,
peng2015genetic}.

Two technological developments of recent years, however,
pose major challenges to the majority of existing simulation methods.
Firstly, fourth-generation sequencing technologies have made
complete chromosome-level assemblies possible~\citep{miga2020telomere},
% Some citations here? Darwin tree of life maybe?
and high quality assemblies are now available for many species.
Thus, modelling genetic variation data as a series of unlinked
non-recombining loci
is no longer a reasonable approximation, and
we must fully account for recombination.
While a genealogical tree relating $n$ samples in the single-locus coalescent
can be simulated in $O(n)$ time~\citep{hudson1990gene},
the coalescent with recombination is far more complex,
and programs such as
Hudson's classical \ms~\citep{hudson2002generating}
can only simulate short segments under the influence of recombination.
The second challenge facing simulation methods is that
sample sizes in genetic studies have grown very quickly
in recent years, enabled by the precipitous fall in genome sequencing costs.
Human datasets like the
UK Biobank~\citep{bycroft2018genome} and
gnomAD~\citep{karczewski2020mutational} now consist of hundreds of
thousands of genomes and many other datasets on a similar scale
are becoming available~\citep{tanjo2021practical}.
Classical simulators such as \ms\ and even fast approximate methods
such as \scrm~\citep{staab2015scrm} simply cannot
cope with such a large number of samples.

@@

% We can simulate so much more now
The 1.0 release of \msprime\ marks a major increase in the
biological realism of simulations
and the breadth of available features, without a loss of efficiency.
These abilities will allow researchers to perform more robust power analyses,
more reliably test new methods,
carry out more reliable inferences,
and explore the properties of more theoretical models.
These advances have only been possible thanks to a distributed,
collaborative model of software development,
and the work of many people.
% We hope the tool continues to be useful to many.

@@

Even though simulation has long been a vital tool in population genetics,
such collaborative software development has historically been uncommon.
A huge proliferation of tools have been published
(the references here are not exhaustive)
and only a small minority of these are actively developed and
maintained today. The ecosystem is highly fragmented, with numerous different
ways of specifying parameters and representing results, and there are
significant software quality issues at all stages. This is unsurprising, since
the majority of simulation software development is performed by students, who
are themselves learning to code, often without formal training in software
development.
Although this ``Haldane's sieve'' of software has produced many good tools
and enabled decades of research,
it also represents a missed opportunity to invest as a community
in shared infrastructure and mentorship in good software development practice.

% THERE IS A BETTER WAY!!!
Scientific software is vital apparatus, and must be engineered
to a high quality if we are to trust its results. There is a growing
realisation across the sciences
% More citations?
\citep[e.g.][]{siepel2019challenges,harris2020array,gardner2021sustained}
that investing in
shared community infrastructure produces better results than a
proliferation of individually maintained tools, allowing scientists
to focus on their specific questions rather than software engineering.
\texttt{Msprime} 1.0 is the result of such a community process,
with features added by motivated users, taking advantage of the
established development practices and infrastructure.
% JK: I'm going out on a limb here - do folks agree?
% PLR- personally I have learned SO MUCH, and I'm passing it on to others
Software development in a welcoming community,
with mentorship by experienced developers,
is a useful experience for many users.
The skills that contributors learn
can lead to greatly increased productivity in subsequent
work (e.g., through more reliable code and better debugging skills).
We hope that users who find that
features they require are missing will continue to contribute to
\msprime, leading to a community project that is both high quality
and sustainable in the long term.

The succinct tree sequence data structure developed
for \msprime\ allows users a view of not only
genetic variation but also the genetic ancestry that produced that variation.
Recent breakthroughs in methods to infer genetic ancestry in recombining
organisms~\citep{kelleher2019inferring,
speidel2019method,wohns2021unified,schaefer2021ancestral,speidel2021inferring}
have made it possible to estimate such ancestry from real data at scale for
the first time~\citep{harris2019database,tang2019genealogy}.
Given such inferred ancestry, many exciting applications
become possible. For example, \cite{osmond2021estimating}
developed a method to estimate the location of genetic ancestors
based on inferred trees, and other uses are sure to follow.
Since the inferred genetic ancestry
becomes the input for other downstream inferences, it is vitally
important that these primary inferences are thoroughly validated,
with the detailed properties of the inferred ancestries
catalogued and understood.
\texttt{Msprime} will continue to be an important tool for these inferences and validations,
and in this context the ability to interoperate
with other methods---particularly forwards simulators---through
the succinct tree sequence data structure and \tskit\ library
will be essential.
% this final sentence is hard to parse

\section*{Acknowledgements}
We would like to thank Iain Mathieson and Alywyn Scally for helpful comments
on the manuscript.

% TODO alphabetise this by author I guess
Jerome Kelleher is supported by the Robertson Foundation.
Jere Koskela is supported in part by EPSRC grant EP/R044732/1.
ADK was supported by NIH awards R01GM117241 and R01HG010774.
Franz Baumdicker is funded by the Deutsche Forschungsgemeinschaft
EXC 2064/1 -- Project number 390727645, and EXC 2124 -- Project number 390838134.
Graham Gower was supported by a Villum Fonden Young Investigator award to Fernando Racimo (project no. 00025300).
BE was supported by  DFG grant  273887127 through Priority
Programme SPP 1819: Rapid Evolutionary Adaptation (grant STE 325/17-2) to Wolfgang Stephan; BE would also like to acknowledge  funding through The Icelandic Research
Centre (Rann{\'i}s) through an Icelandic Research Fund Grant of  Excellence nr.\ 185151-051 to Einar \'Arnason,  Katr\'in Halld\'orsd\'ottir,  Alison Etheridge,  Wolfgang Stephan, and BE.
Gregor Gorjanc is supported by the Chancellor's Fellowship of the University of
Edinburgh and the BBSRC grant to The Roslin Institute BBS/E/D/30002275.
SG acknowledges funding from the Canada Research Chairs Program, from the
Canadian Institutes of Health Research PJT 173300, and from the Canadian
Foundation for Innovation.

>>>>

\begin{abstract}
Stochastic simulation is a key tool in population genetics,
since the models involved are often analytically intractable
and simulation is usually the only way of obtaining ground-truth data to
evaluate inferences.
Because of this necessity, a large number
of specialised simulation programs have been developed, each
filling a particular niche, but with largely overlapping functionality
and a substantial duplication of effort.
Here, we introduce \msprime\ version 1.0, which
efficiently implements ancestry and mutation simulations based on
the succinct tree sequence data structure and \tskit\ library.
We summarise \msprime's many features, and show that
its performance is excellent, often many times faster
and more memory efficient than specialised alternatives.
These high-performance features have been thoroughly
tested and validated, and built using a collaborative,
open source development model,
which reduces duplication of effort and promotes
software quality via community engagement.
% "Bazaar" development model: https://en.wikipedia.org/wiki/The_Cathedral_and_the_Bazaar
\end{abstract}

@@

The coalescent
process~\citep{kingman1982coalescent,hudson1983testing,tajima1983evolutionary}
models the ancestry of a set of sampled genomes,
providing a mathematical description of the
genealogical tree that relates the samples to one another.
It has proved to be a powerful model,
and is now central to
population genetics~\citep{hudson1990gene,hein2004gene,wakely2008coalescent}.
The coalescent is a very efficient framework for population genetic
simulation, because it allows us to simulate the genetic ancestry for
a sample from an idealised population model, without explicitly representing
the population in memory or stepping through the generations.
Indeed, \cite{hudson1983testing} independently derived the
% Anyone disagree here? That's the impression I get from the paper.
coalescent \emph{in order to} efficiently simulate data,
and used these simulations to characterise an analytically
intractable distribution.
This inherent efficiency, and the great utility of simulations for a wide
range of purposes, has led to dozens of different tools
being developed over the decades~\citep{carvajal2008simulation,liu2008survey,
arenas2012simulation,yuan2012overview,hoban2012computer,yang2014critical,
peng2015genetic}.

Two technological developments of recent years, however,
pose major challenges to most existing simulation methods.
Firstly, fourth-generation sequencing technologies have made
complete chromosome-level assemblies possible~\citep{miga2020telomere},
% Some citations here? Darwin tree of life maybe?
and high quality assemblies are now available for many species.
Thus, modelling genetic variation data as a series of unlinked
non-recombining loci
is no longer a reasonable approximation, and
we must fully account for recombination.
However, while a genealogical tree relating $n$ samples in the single-locus coalescent
can be simulated in $O(n)$ time~\citep{hudson1990gene},
the coalescent with recombination is far more complex,
and programs such as
Hudson's classical \ms~\citep{hudson2002generating}
can only simulate short segments under the influence of recombination.
The second challenge facing simulation methods is that
sample sizes in genetic studies have grown very quickly
in recent years, enabled by the precipitous fall in genome sequencing costs.
Human datasets like the
UK Biobank~\citep{bycroft2018genome} and
gnomAD~\citep{karczewski2020mutational} now consist of hundreds of
thousands of genomes and many other datasets on a similar scale
are becoming available~\citep{tanjo2021practical}.
Classical simulators such as \ms\ and even fast approximate methods
such as \scrm~\citep{staab2015scrm} simply cannot
cope with such a large number of samples.

@@

% We can simulate so much more now
The 1.0 release of \msprime\ marks a major increase in the
breadth of available features and the
potential biological realism of simulations.
These abilities will allow researchers to perform more robust power analyses,
more reliably test new methods,
carry out more reliable inferences,
and more throughly explore the properties of theoretical models.
Despite this complexity and generality,
\msprime's performance is state-of-the-art
and all features are thoroughly tested and statistically validated.
These advances have only been possible thanks to a distributed,
collaborative model of software development,
and the work of many people.
% We hope the tool continues to be useful to many.

@@

Even though simulation has long been a vital tool in population genetics,
such collaborative software development has historically been uncommon.
A huge proliferation of tools have been published
(the references here are not exhaustive)
and only a small minority of these are actively developed and
maintained today. The ecosystem is highly fragmented, with numerous different
ways of specifying parameters and representing results, and there are
significant software quality issues at all stages. This is unsurprising, since
the majority of simulation software development is performed by students,
often without formal training in software development.
Although this ``Haldane's sieve'' of software has produced many good tools
and enabled decades of research,
it also represents a missed opportunity to invest as a community
in shared infrastructure and mentorship in good software development practice.

% THERE IS A BETTER WAY!!!
Scientific software is vital apparatus, and must be engineered
to a high quality if we are to trust its results. There is a growing
realisation across the sciences
% More citations?
\citep[e.g.][]{siepel2019challenges,harris2020array,gardner2021sustained}
that investing in
shared community infrastructure produces better results than a
proliferation of individually maintained tools, allowing scientists
to focus on their specific questions rather than software engineering.
\texttt{Msprime} 1.0 is the result of such a community process,
with features added by motivated users, taking advantage of the
established development practices and infrastructure.
Software development in a welcoming community,
with mentorship by experienced developers,
is a useful experience for many users.
The skills that contributors learn
can lead to greatly increased productivity in subsequent
work (e.g., through more reliable code and better debugging skills).
We hope that users who find that
features they require are missing will continue to contribute to
\msprime, leading to a community project that is both high quality
and sustainable in the long term.

The succinct tree sequence data structure developed
for \msprime\ provides a view of not only
genetic variation, but also the genetic ancestry that produced that variation.
Recent breakthroughs in methods to infer genetic ancestry in recombining
organisms~\citep{rasmussen2014genome,kelleher2019inferring,
speidel2019method,wohns2021unified,schaefer2021ancestral,speidel2021inferring}
have made it possible to estimate such ancestry from real data at scale for
the first time~\citep{harris2019database,tang2019genealogy}.
Given such inferred ancestry, many exciting applications
become possible. For example, \cite{osmond2021estimating}
developed a method to estimate the location of genetic ancestors
based on inferred trees, and other uses are sure to follow.
Since the inferred genetic ancestry
becomes the input for other downstream inferences, it is vitally
important that these primary inferences are thoroughly validated,
with the detailed properties of the inferred ancestries
catalogued and understood.
\texttt{Msprime} will continue to be an important tool for these inferences and validations,
and in this context the ability to interoperate
with other methods---particularly forwards simulators---through
the succinct tree sequence data structure and \tskit\ library
will be essential.
% this final sentence is hard to parse

\section*{Acknowledgements}
We would like to thank Iain Mathieson and Alywyn Scally for helpful comments
on the manuscript.
% alphabetised by author. Names spelled out if there's an initial collision
ADK was supported by NIH awards R01GM117241 and R01HG010774.
BE was supported by  DFG grant  273887127 through Priority Programme SPP 1819: Rapid Evolutionary Adaptation (grant STE 325/17-2) to Wolfgang Stephan; BE would also like to acknowledge  funding through The Icelandic Research Centre (Rann{\'i}s) through an Icelandic Research Fund Grant of  Excellence nr.\ 185151-051 to Einar \'Arnason,  Katr\'in Halld\'orsd\'ottir,  Alison Etheridge,  Wolfgang Stephan, and BE.
FB is funded by the Deutsche Forschungsgemeinschaft EXC 2064/1 -- Project number 390727645, and EXC 2124 -- Project number 390838134.
Graham Gower was supported by a Villum Fonden Young Investigator award to Fernando Racimo (project no. 00025300).
Gregor Gorjanc is supported by the Chancellor's Fellowship of the University of Edinburgh and the BBSRC grant to The Roslin Institute BBS/E/D/30002275.
Jere Koskela is supported in part by EPSRC grant EP/R044732/1.
Jerome Kelleher is supported by the Robertson Foundation.
SG acknowledges funding from the Canada Research Chairs Program, from the Canadian Institutes of Health Research PJT 173300, and from the Canadian Foundation for Innovation.
